From Reversible Programs to \\ Univalent Universes and Back

### 2021-07-09 Pi+/paper/equivalence.tex

\section{Going from \texorpdfstring{$\PiLang$}{Pi} to \texorpdfstring{$\UFin$}{UFin} and back}
\label{sec:equivalence}

@@

In this section, we consider the additive fragment $\PiPlusLang$ of $\PiLang$. This is the language that we interpret to
$\UFin$ and back, using the tools developed in the previous sections. Further, we go through an intermediate step of the
language $\PiHatLang$, which is a simplified variant of $\PiPlusLang$ that uses adjacent transpositions for combinators,
while preserving all the required structure.

We present the types and 1-combinators of $\PiPlusLang$ and $\PiHatLang$ in~\Cref*{fig:piplus,fig:pihat} respectively,
eliding the 2-combinators for brevity. \note{We enforce that there is a unique 2-combinator between compatible
1-combinators, by introducing a level 3, which is trivial in the same way as in the case of $\PiLang$, as described
before.} They can be found in the appendix and in the accompanying Agda code.

@@

To state our results formally, we organise the syntax for each language using a technical device, called a syntactic
category. Since the languages have higher-dimensional structure, the syntactic categories here are actually
(locally-strict) weak 2-categories. We define them formally in the appendix, and only state our results here.

For each of the $\PiLang$, $\PiPlusLang$ and $\PiHatLang$ languages, their syntactic categories, respectively $\PiCat$,
$\PiPlusCat$ and $\PiHatCat$, have 0-cells for types, 1-cells for 1-combinators, and 2-cells for 2-combinators. We can
show that these are $(2,0)$-categories, since all the 1-cells and 2-cells are invertible. They are also locally-strict,
or locally-posetal, because there is at most one 2-cell between compatible 1-cells.

@@

\begin{proposition}
  There are two symmetric monoidal structures on $\PiPlusCat$, given by $(\zerot, +)$ and $(\onet, \times)$, with
  $\times$ distributing over $+$, giving it a rig structure, as described in \todo{cite earlier section}.
\end{proposition}
\todo{either prove it or cite something}

@@

\begin{proposition}
  $\evalt$ gives a rig biequivalence between $\PiCat$ and $\PiPlusCat$.
\end{proposition}

@@

\begin{proposition}
  $\evalp/\quotep$ give a symmetric monoidal biequivalence between $\PiPlusCat$ and $\PiHatCat$.
\end{proposition}

@@

\begin{proposition}
  $\evalh/\quoteh$ give a symmetric monoidal biequivalence between $\PiHatCat$ and $\UFin$.
\end{proposition}

>>>>

\section{Correspondence between \texorpdfstring{$\PiLang$}{Pi} and \texorpdfstring{$\UFin$}{UFin}}
\label{sec:equivalence}

@@

In this section, we first translate $\PiLang$ to its additive fragment $\PiPlusLang$. This is the language that we
interpret to $\UFin$ and back, using the tools developed in the previous sections. Further, we go through an
intermediate step of the language $\PiHatLang$, which is a simplified variant of $\PiPlusLang$ that uses adjacent
transpositions for combinators, while preserving all the required structure.

We present the types and 1-combinators of $\PiPlusLang$ and $\PiHatLang$ in~\Cref*{fig:piplus,fig:pihat} respectively,
eliding the 2-combinators for brevity. We enforce that there is a unique 2-combinator between compatible 1-combinators,
by relating them with a truncation. These can be found in the appendix and in the accompanying Agda code
(~\cref{app:level-two}).

@@

To state our results formally, we organise the syntax for each language using a technical device, called a syntactic
category. We define them formally in the appendix, and only state our results here. For each of the $\PiLang$,
$\PiPlusLang$ and $\PiHatLang$ languages, their syntactic categories, respectively $\PiCat$, $\PiPlusCat$ and
$\PiHatCat$, have 0-cells for types, 1-cells for 1-combinators, and 2-cells for 2-combinators. We can shoe that these
syntactic categories here are actually $(2,0)$-categories, since all the 1-cells and 2-cells are invertible. They are
also locally-strict, or locally-posetal, because there is at most one 2-cell between compatible 1-cells.

@@

\begin{proposition}
  There are two symmetric monoidal structures on $\PiPlusCat$, given by $(\zerot, +)$ and $(\onet, \times)$, with
  $\times$ distributing over $+$, giving it a rig structure.
\end{proposition}

@@

\begin{proposition}
  $\evalt$ gives a rig equivalence between $\PiCat$ and $\PiPlusCat$.
\end{proposition}

@@

\begin{proposition}
  $\evalp/\quotep$ give a symmetric monoidal equivalence between $\PiPlusCat$ and $\PiHatCat$.
\end{proposition}

@@

\begin{proposition}
  $\evalh/\quoteh$ give a symmetric monoidal equivalence between $\PiHatCat$ and $\UFin$.
\end{proposition}
